---
title: "My Fault - Anyflights Package Doesn't Have Geographic Data"
output: html_document
date: "2024-04-22"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Was there a time when you felt like a genius only to realize you had overlooked something obvious? In this project, I initially believed that the 'anyflights' package did not include longitude and latitude information for airports. Consequently, I imported the 'us_map' package to extract geographical information about each state. However, I encountered an issue when merging my datasets, as 'us_map' provided information at the state level rather than for individual airports. To address this, I calculated the average longitude and latitude for each state and matched airports to their corresponding states. This allowed me to successfully merge the datasets and create a visualization of flight routes using ggplot. I felt like a genius! That is, until I discovered that the 'anyflights' package did indeed include longitude and latitude information for each airport.

```{r cars}
library("anyflights")
library("dplyr")

data <- read.csv("combined_flights_weather.csv")

data <- data %>%
  filter(year.x == 2021) %>% # This is a switch to see the flights traces from any year interesting
  select(origin, dest)


library(maps)

library(dplyr)
library(tidyr)
library(ggplot2)

airport_states <- data.frame(
  airport = c("JFK","LAX","RSW", "SJU", "BQN", "FLL", "BUF", "MCO", "MIA", "ATL", "ORD", "DFW","DEN", "CLT", "SFO", "SEA", "PHX", "BOS", "SLC", "SYR", "CHS","LAS", "MSP", "ROC", "PSP","SRQ", "TPA", "MSY", "RDU", "JAX","SNA", "PBI", "SJC", "IAD", "SAN", "SAV", "STT", "DAB", "LGB", "BWI","DTW", "PDX", "PWM", "HNL", "BUR", "EGE", "BTV", "DCA", "AUS", "ORF",
              "IND", "BNA", "HOU", "CLE", "CVG", "RIC", "PIT", "CMH", "SAT", "SMF",
              "ONT", "RNO", "OAK", "PSE", "ABQ", "ORH", "JAC", "IAH", "HDN", "ACK",
              "BZN", "EYW", "EWR", "PHL", "DAL", "ASE", "MRY", "OGG", "MDW", "LIH",
              "COS", "MFR", "SBP", "KOA", "TUS", "ACV", "FAT", "STL", "RDM", "SBA",
              "ELP", "SGF", "SUN", "MMH", "BOI", "SGU", "ITO", "OMA", 
              "MKE","MEM", "PRC", "GEG", "BDL", "ANC", "MTJ", "PVU", "TUL", "OKC",
              "STS", "RDD", "MSN", "PSC", "MSO", "GRR", "PAE"),
  state = c("new york", "california", "florida", "puerto rico", "puerto rico", "california", "new york","florida", "florida", "georgia", "illinois", "texas","colorado", 
            "north carolina", "california", "washington", "arizona", "massachusetts", 
            "utah", "new york", "south carolina", "nevada", "louisiana", "new york", 
            "california", 
            "florida", "florida","louisiana", "north carolina","florida",
            "california", "wyoming", "california", "virginia", "california", "georgia", "virgin islands", "florida", "california", "maryland","michigan", "oregon", "maine", "hawaii", "california", "colorado", "vermont", "virginia", "texas", "virginia",
            "indiana", "tennessee","texas", "ohio", "kentucky", "virginia", "pennsylvania", "ohio", "texas","california",
            "california", "nevada", "california", "puerto rico", "new mexico", "massachusetts", "wyoming", "florida", "new jersey", "massachusetts",
            "montana", "florida", "new jersey","pensylvania", "texas", "colorado", "california", "hawaii", "illinois", "hawaii",
            "colorado", "oregon", "california", "hawaii", "arizona","california", "california", "missouri", "oregon", "california",
            "texas", "missouri", "idaho", "california","idaho", "utah", "hawaii", "nebraska", 
            "wisconsin", "tennessee","arizona","washington","connecticut","alaska","colorado","utah", "oklahoma","oklahoma",
            "california", "california", "wisconsin", "washington", "montana", "michigan", 
            "washington")
)

nrow(airport_states$airport) == nrow(airport_states$state)

data$dest <- airport_states$state[match(data$dest, airport_states$airport)]



# 1. Merge the average latitude and longitude for each state into our data table
states <- map_data("state")

us_state_coords <- states %>%
  group_by(region) %>%
  summarize(avg_long = mean(long),
            avg_lat = mean(lat))


# 2. Merge the latitude and longitude of states with "data" using state names

airport_states <- airport_states %>%
  mutate(state = tolower(state))

state_coords_merged <- inner_join(us_state_coords, airport_states, by = c("region" = "state"))
 
data_with_coords <- inner_join(data, state_coords_merged, by = c("dest" = "region"))


# 3. Merge the latitude and longitude information for JFK and LAX into the existing table as well
lax_coords <- us_state_coords %>%
  filter(region == "california") %>%
  select(avg_long, avg_lat)

jfk_coords <- us_state_coords %>%
  filter(region == "new york") %>%
  select(avg_long, avg_lat)

state_coords_merged <- data_with_coords %>%
  mutate(jfk_long = ifelse(origin == "JFK", jfk_coords$avg_long, NA),
         jfk_lat = ifelse(origin == "JFK", jfk_coords$avg_lat, NA),
         lax_long = ifelse(origin == "LAX", lax_coords$avg_long, NA),
         lax_lat = ifelse(origin == "LAX", lax_coords$avg_lat, NA))


# 4. Plotting flight routes from LAX and JFK airports to airports nationwide
us_states <- map_data("state")

jfk_data <- data.frame(region = "new york", avg_long = jfk_coords$avg_long, avg_lat = jfk_coords$avg_lat)
lax_data <- data.frame(region = "california", avg_long = lax_coords$avg_long, avg_lat = lax_coords$avg_lat)

us_states <- map_data("state")

jfk_data <- data.frame(region = "new york", avg_long = jfk_coords$avg_long, avg_lat = jfk_coords$avg_lat)
lax_data <- data.frame(region = "california", avg_long = lax_coords$avg_long, avg_lat = lax_coords$avg_lat)

segment_counts <- data.frame(Airport = c("JFK", "LAX"),
                             Count = c(sum(!is.na(state_coords_merged$jfk_long)),
                                       sum(!is.na(state_coords_merged$lax_long))))

ggplot(state_coords_merged, aes(x = avg_long, y = avg_lat)) +
  geom_polygon(data = us_states, aes(x = long, y = lat, group = group), fill = "grey90", color = "black") +
  geom_point() +
  geom_segment(aes(x = jfk_long, y = jfk_lat, xend = avg_long, yend = avg_lat), color = "#00AFBB") +  # Draw segments connecting to JFK
  geom_segment(aes(x = lax_long, y = lax_lat, xend = avg_long, yend = avg_lat), color = "#C4961A") +  # Draw segments connecting to LAX
  geom_text(data = jfk_data, aes(x = avg_long, y = avg_lat, label = "JFK"), color = "#00AFBB", size = 3, vjust = -0.5) +
  geom_text(data = lax_data, aes(x = avg_long, y = avg_lat, label = "LAX"), color = "#C4961A", size = 3, vjust = -0.5) +
  labs(title = "Flight Routes from LAX and JFK Airports to Airports Across the Country (2021)",
       x = "Longitude", y = "Latitude",
       color = "Origin Airport",
       subtitle = "Number of flights:",
       caption = paste("JFK:", segment_counts$Count[1], "\n", "LAX:", segment_counts$Count[2])) +
  theme_minimal()

```

